\documentclass[phd,cisa]{infthesis}

\usepackage{proof}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{subfigure}
\usepackage{float}
\usepackage{url}
\usepackage{multirow}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{color}
\usepackage{fancybox}

\definecolor{gray}{rgb}{0.9,0.9,0.9}

\allowdisplaybreaks
\floatstyle{boxed}
\newfloat{boxedfigure}{H}{lobf}
\floatname{boxedfigure}{Figure}
\restylefloat{plain}
\makeatletter
\let\c@boxedfigure\c@figure
\makeatother 
\renewcommand{\theboxedfigure}{\arabic{chapter}.\arabic{figure}}

\newlength{\boxwidth}
\setlength{\boxwidth}{\linewidth}
\addtolength{\boxwidth}{3ex}

\newtheorem*{proposition}{Proposition}
\newtheorem*{lemma}{Lemma}

\newcommand{\bind}{\vartriangleright}
\newcommand{\code}{\texttt}
\newcommand{\adjoin}[2]{\{#1\}\,\cup\,#2}
\newcommand{\cons}[2]{#1\,::\,#2}
\newcommand{\append}[2]{#1 + #2}
\newcommand{\el}[2]{\code{el}\ #1\ #2}
\renewcommand{\between}[3]{\code{between}\ #1\ #2\ #3}
\newcommand{\online}[2]{\code{on\_line}\ #1\ #2}
\newcommand{\onplane}[2]{\code{on\_plane}\ #1\ #2}
\newcommand{\onpolypath}[2]{\code{on\_polypath}\ #1\ #2}
\newcommand{\collinearthree}[3]{($\exists$a.\; #1 on a $\wedge$ #2 on a $\wedge$ #3 on a)}
\newcommand{\collinearfour}[4]{($\exists$a.\; #1 on a $\wedge$ #2 on a $\wedge$ #3 on a $\wedge$ #4 on a)}
\newcommand{\collinearfive}[5]{($\exists$a.\; #1 on a $\wedge$ #2 on a $\wedge$ #3 on a $\wedge$ #4 on a $\wedge$ #5 on a)}
\newcommand{\planarthree}[3]{($\exists\alpha$.\; #1 on $\alpha$ $\wedge$ #2 on $\alpha$ $\wedge$ #3 on a)}
\newcommand{\planarfour}[4]{($\exists\alpha$.\; #1 on $\alpha$ $\wedge$ #2 on $\alpha$ $\wedge$ #3 on $\alpha$ $\wedge$ #4 on a)}
\newcommand{\planarfive}[5]{($\exists\alpha$.\; #1 on $\alpha$ $\wedge$ #2 on $\alpha$ $\wedge$ #3 on $\alpha$ $\wedge$ #4 on $\alpha$ $\wedge$ #5 on a)}
\newcommand{\planarsix}[6]{($\exists\alpha$.\; #1 on $\alpha$ $\wedge$ #2 on $\alpha$ $\wedge$ #3 on $\alpha$ $\wedge$ #4 on $\alpha$ $\wedge$ #5 on a $\wedge$ #6 on a)}
\newcommand{\planarseven}[7]{($\exists\alpha$.\; #1 on $\alpha$ $\wedge$ #2 on $\alpha$ $\wedge$ #3 on $\alpha$ $\wedge$ #4 on $\alpha$ $\wedge$ #5 on a $\wedge$ #6 on a $\wedge$ #7 on a)}
\newcommand{\ttriangle}[3]{$\neg$($\exists$a.\; #1 on a $\wedge$ #2 on a $\wedge$ #3 on a)}

\newcommand{\Triangle}[4]{\neg(\exists #1.\; \online{#2}{#1}\wedge \online{#3}{#1}\wedge \online{#4}{#1})}
\newcommand{\fourbet}[4]{\between{#1}{#2}{#3}\wedge\between{#1}{#3}{#4}}
\newcommand{\fourbetlong}[4]{\between{#1}{#2}{#3}\wedge\between{#1}{#2}{#4}\wedge\between{#1}{#3}{#4}\wedge\between{#2}{#3}{#4}}
\newcommand{\bounds}[3]{\text{bounds}\,#1\,#2\,#3}

\title{Ordered Geometry in Hilbert's \emph{Grundlagen der Geometrie}}
\author{Phil Scott}

\abstract{The \emph{Grundlagen der Geometrie} brought Euclid's ancient axioms up to the standards of modern logic, anticipating a completely mechanical verification of their theorems. There are five groups of axioms, each focused on a logical feature of Euclidean geometry. The first two groups give us \emph{ordered geometry}, a highly limited setting where there is no talk of measure or angle. From these, we mechanically verify the Polygonal Jordan Curve Theorem, a result of much generality given the setting, and subtle enough to warrant a full verification.

Along the way, we describe and implement a general-purpose algebraic language for proof search, which we use to automate arguments from the first axiom group. We then follow Hilbert through the preliminary definitions and theorems that lead up to his statement of the Polygonal Jordan Curve Theorem. These, once formalised and verified, give us a final piece of automation. Suitably armed, we can then tackle the main theorem.
}
\begin{document}

%% First, the preliminary pages
\begin{preliminary}

%% This creates the title page
\maketitle

%% Acknowledgements
\begin{acknowledgements}
Many thanks go to my supervisor Jacques Fleuriot for his support and encouragement. Thanks to Laura Meikle for spotting a simplification of Theorem~\ref{eq:PaschPointSet}, and especially to Steven Obua for helping me prove transitivity of polygonal rotations described in \S\ref{sec:PolygonRotation}.
\end{acknowledgements}

%% Next we need to have the declaration.
\begin{declaration}
  I declare that this thesis was composed by myself, that the work contained herein is my own except for work included which forms part of jointly-authored publications. Our contribution and that of the other authors to this work is explicitly indicated below. We confirm that appropriate credit has been given within the thesis where reference has been made to the work of others, and that this work has not been submitted for any other degree or professional qualification except as specified.

In Chapters~\ref{chapter:Automation} and~\ref{chapter:Group2Eval}, we expand on a combinator language which evolved between \emph{Composable Discovery Engines for Interactive Theorem Proving} and \emph{A Combinator Language for Theorem Discovery}, published respectively in \emph{Interactive Theorem Proving} in 2011 and \emph{Intelligent Computer Mathematics} in 2012. The intended application of this language was first described in \emph{An Investigation of Hilbert's Implicit Reasoning through Proof
Discovery in Idle-Time}, published in {\em Automated Deduction in Geometry} in 2010, and our analysis from this paper has been updated and can be found in~\S\ref{sec:FourAlternative} and~\S\ref{sec:FindingAEH}. The work was co-authored with Jacques Fleuriot.
   \par
   \vspace{1in}\raggedleft({\em Phil Scott})
 \end{declaration}

%% Create the table of contents
\tableofcontents

\end{preliminary}

\parindent 0pt\parskip 0.5ex

\include{introduction/introduction}

\include{background/background}

\include{axioms/axioms}

\include{automation/automation}

\include{group2eval/group2eval}

\include{linearOrder/linearOrder}

\include{halfPlanes/halfPlanes}

\include{jordan/jordan}
\include{jordanFormalisation/jordanFormalisation}
\include{jordanVerification1/jordanVerification}
\include{jordanVerification2/jordanVerification}
\include{conclusion/conclusion}

\bibliographystyle{plain}
\bibliography{formalising}

\appendix
%\include{appendices/group1App}
\include{appendices/group2App}
\include{appendices/jordanVerificationApp}
\include{appendices/jordanVerificationExtra}

\end{document}

